---
title: "Description of the mastitis model"
author: "Aurélien Madouasse"
date: "2022-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(DiagrammeR)
```

# Objective

The objective of this document is to describe the mastitis model and its translation into code.

# Model assumptions

## Three state model

Considering individual cows at a point in time, we define 3 states with regard to mastitis.

1. Healthy (H)
2. Contagious mastitis (C)
3. Environmental mastitis (E)

Monthly transition between these states is assumed to be Markovian.

```{r dag mastitis 3 states, echo = FALSE}
grViz("
 digraph mastitis_3_states{
 
 	  # Nodes
	  node [shape = circle]
	  H [label = 'H']
	  C [label = 'C']
	  E [label = 'E']
	  
	  H -> C
	  H -> E
	  H -> H
	  
	  C -> H
	  C -> E
	  C -> C
	  
	  E -> H
	  E -> C
	  E -> E
 }
 ")
```

The assumption of the Markovian transition is that the *state* at time $t+1$ only depends on the state at time $t$. The probabilities of transition can be described with a square matrix with the states at time $t$ in rows and the states at time $t+1$ in columns. The diagonal of this matrix represent the probabilities of remaining in the same state. From an initial state, the sum of the probabilities of reaching all states has to be one. 

We can construct the transition matrix on an arbitrary scale and then scale it in order for the sum of the values on a given row to be one.

```{r}
A <- matrix(
  c(1,   .1, .1, 
    .01,  1, .01, 
    100, .1,  1),
  byrow = TRUE,
  ncol = 3,
  dimnames = list(c("H", "C", "E"), c("H", "C", "E"))
)

A
```

The matrix is rescaled.

```{r}
for(i in 1:3){
  
  A[i,] <- A[i,] / sum(A[i,])
  
}

A
```

## Four state model

In this model, the a cow can have an infection with both contagious and environmental pathogens at the same time. 

```{r dag mastitis 4 states, echo = FALSE}
grViz("
 digraph mastitis_4_states{
 
 	  # Nodes
	  node [shape = circle]
	  H [label = 'H']
	  C [label = 'C=1, E=0']
	  E [label = 'C=0, E=1']
	  CE [label = 'C=1, E=1']
	  
	  H -> C
	  H -> E
	  H -> H
	  H -> CE
	  
	  C -> H
	  C -> E
	  C -> C
	  C -> CE
	  
	  E -> H
	  E -> C
	  E -> E
	  E -> CE
	  
	  CE -> H
	  CE -> C
	  CE -> E
	  CE -> CE

 }
 ")
```

There are now 16 possible transitions, defined by a 4x4 matrix. The additional state represents the probability of having both contagious and environmental mastitis at the same time. This matrix can be represented as follows:

```{r}
B <- matrix(
  paste0(rep(paste0("b", 1:4), each=4), 1:4),
  byrow = TRUE,
  ncol = 4,
  dimnames = list(c("H", "C", "E", "CE"), c("H", "C", "E", "CE"))
)
B
```

# Constructing the transition matrix for the 4 state model

The transition matrix for the 4 states is built from the transition matrix for the 3 state model. Transitions from environmental to contagious and from contagious to environmental are supposed impossible. This will prevent the model from assuming that cows with a high SCC between 2 consecutive months jump from environmental to contagious and vice versa.

```{r}
B <- matrix(
  rep(0, 16),
  byrow = TRUE,
  ncol = 4,
  dimnames = list(c("H", "C", "E", "CE"), c("H", "C", "E", "CE"))
)

## From healthy
B[1, 1] <- A[1, 1]
# this is to keep the sum of the probabilities on this row at 1
# while conserving the proportions in the different categories
factr <- (A[1, 2] + A[1, 3]) / (A[1, 2] + A[1, 3] + A[1, 2] * A[1, 3])

B[1, 2] <- A[1, 2] * factr
B[1, 3] <- A[1, 3]  * factr
B[1, 4] <- prod(A[1, 2:3])  * factr

## From contagious
# transitions to environmental and environmental + contagious are impossible
denom <- sum(A[2, 1:2])

B[2, 1] <- A[2, 1] / denom
B[2, 2] <- A[2, 2] / denom
B[2, 3] <- 0
B[2, 4] <- 0

## From environmental
# transitions to contagious and environmental + contagious are impossible
denom <- sum(A[3, c(1, 3)])
B[3, 1] <- A[3, 1] / denom
B[3, 2] <- 0
B[3, 3] <- A[3, 3] / denom
B[3, 4] <- 0

## From contagious + environmental
# constraint to ensure that the row sums to one
fctr <- 1 / (A[2, 1] + A[3, 1] + A[2, 1] * A[3, 1])

B[4, 1] <- A[2, 1] * A[3, 1] * fctr
B[4, 2] <- A[3, 1] * fctr
B[4, 3] <- A[2, 1] * fctr
B[4, 4] <- 1 - sum(B[4, 1:3])

rm(list = c("denom", "factr"))

B
```




